---
title: 'Reproducible Research: Peer Assessment 2'
author: "Jesse Shaw"
date: "March 22, 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#Dangerous Weather Events- Mortality and Economic Impact 

The U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database makes 
public characteristics of severe storms and extreme weather events.  The data spans 1950
to 2011 making it possible to analyze which storm categories cause the most economic damage
as well as injuries and fatalities.  Because earlier recorded data is incomplete, the scope 
of this analysis is limited to the last 20 years of published data (1991-2011).
 
 The data dictionary can be found here:
 https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
 
##Load Required Libraries
```{r libraries, echo=TRUE}
library(R.utils)
```

##Data Processing
```{r data_load, echo=TRUE}
##Set File URL
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

##Check For File Prior To Download
if (!"stormData.csv.bz2" %in% dir("./")) {
        download.file(fileURL, destfile = "stormData.csv.bz2", method="curl")
        bunzip2("stormData.csv.bz2", overwrite=T, remove=F)
} 

##Read StormData File
stormData <- read.csv("stormData.csv", sep = ",")
```

A portion of the original dataset is as follows:
```{r first_output, echo=TRUE}
print(stormData[1:5,])
```

Since we are only interested in property damage, injury, and fatalities, the following fields will be used:

* FATALITIES
* INJURIES
* PROPDMG
* PROPDMGEXP
* CROPDMG
* CROPDMGEXP

```{r data_processing, echo=TRUE}
##Subset Original Dataset
eventData <- stormData[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

#Property Damage Preparation
eventData$PROPEXP[eventData$PROPDMGEXP == "K"] <- 1000
eventData$PROPEXP[eventData$PROPDMGEXP == "M"] <- 1e+06
eventData$PROPEXP[eventData$PROPDMGEXP == ""]  <- 1
eventData$PROPEXP[eventData$PROPDMGEXP == "B"] <- 1e+09
eventData$PROPEXP[eventData$PROPDMGEXP == "m"] <- 1e+06
eventData$PROPEXP[eventData$PROPDMGEXP == "0"] <- 1
eventData$PROPEXP[eventData$PROPDMGEXP == "5"] <- 1e+05
eventData$PROPEXP[eventData$PROPDMGEXP == "6"] <- 1e+06
eventData$PROPEXP[eventData$PROPDMGEXP == "4"] <- 10000
eventData$PROPEXP[eventData$PROPDMGEXP == "2"] <- 100
eventData$PROPEXP[eventData$PROPDMGEXP == "3"] <- 1000
eventData$PROPEXP[eventData$PROPDMGEXP == "h"] <- 100
eventData$PROPEXP[eventData$PROPDMGEXP == "7"] <- 1e+07
eventData$PROPEXP[eventData$PROPDMGEXP == "H"] <- 100
eventData$PROPEXP[eventData$PROPDMGEXP == "1"] <- 10
eventData$PROPEXP[eventData$PROPDMGEXP == "8"] <- 1e+08

#Remove Bad Data Influence
eventData$PROPEXP[eventData$PROPDMGEXP == "+"] <- 0
eventData$PROPEXP[eventData$PROPDMGEXP == "-"] <- 0
eventData$PROPEXP[eventData$PROPDMGEXP == "?"] <- 0

#Calculate Total Property Damage
eventData$PROPDMGVAL <- eventData$PROPDMG * eventData$PROPEXP

#Crop Damage Preparation
eventData$CROPEXP[eventData$CROPDMGEXP == "M"] <- 1e+06
eventData$CROPEXP[eventData$CROPDMGEXP == "K"] <- 1000
eventData$CROPEXP[eventData$CROPDMGEXP == "m"] <- 1e+06
eventData$CROPEXP[eventData$CROPDMGEXP == "B"] <- 1e+09
eventData$CROPEXP[eventData$CROPDMGEXP == "0"] <- 1
eventData$CROPEXP[eventData$CROPDMGEXP == "k"] <- 1000
eventData$CROPEXP[eventData$CROPDMGEXP == "2"] <- 100
eventData$CROPEXP[eventData$CROPDMGEXP == ""]  <- 1

#Remove Bad Data Influence
eventData$CROPEXP[eventData$CROPDMGEXP == "?"] <- 0

#Calculate Total Crop Damage
eventData$CROPDMGVAL <- eventData$CROPDMG * eventData$CROPEXP

#Aggregate By Target Events
fatality <- aggregate(FATALITIES ~ EVTYPE, data = eventData, FUN = sum)
top_fatalities <- fatality[order(-fatality$FATALITIES), ][1:10, ]

injury <- aggregate(INJURIES ~ EVTYPE, data = eventData, FUN = sum)
top_injuries <- injury[order(-injury$INJURIES), ][1:10, ]

property <- aggregate(PROPDMGVAL ~ EVTYPE, data = eventData, FUN = sum)
top_property <- property[order(-property$PROPDMGVAL), ][1:10, ]

crop <- aggregate(CROPDMGVAL ~ EVTYPE, data = eventData, FUN = sum)
top_crop <- crop[order(-crop$CROPDMGVAL), ][1:10, ]
```

##Results

* The biggest risk to persons in the United States is tornados.
* The largest contributor to property damage is flooding.
* Water is necessary for crops, but severe fluctuations (drought/flood) in water are the largest contributors to crop damage.

```{r plots, echo=TRUE}
##Plots
par(mfrow=c(2, 2))
IV <- barplot(top_fatalities$FATALITIES, names = top_fatalities$EVTYPE, main = "Fatalities", xaxt="n", space=1)
text(cex=1, x=IV+.25 , y=-4.25 , top_fatalities$EVTYPE , xpd=T , srt=60 , pos=2)

I <- barplot(top_injuries$INJURIES, names = top_injuries$EVTYPE, main= "Injuries", xaxt="n", space=1)
text(cex=1, x=I+.25 , y=-4.25 , top_injuries$EVTYPE , xpd=T , srt=60 , pos=2)

II <- barplot(top_property$PROPDMG, names = top_property$EVTYPE, main="Property Damage", xaxt="n", space=1)
text(cex=1, x=II+.25 , y=-4.25 , top_property$EVTYPE , xpd=T , srt=60 , pos=2)

III <- barplot(top_crop$CROPDMG, names = top_crop$EVTYPE, main="Crop Damage", xaxt="n", space=1)
text(cex=1, x=III+.25 , y=-4.25 , top_crop$EVTYPE , xpd=T , srt=60 , pos=2)
```
